\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}

\begin{document}

\title{Reinforcement Learning in Quantum-Inspired Latent Space}
\author{RL-world-model Overview}
\date{}
\maketitle

\section*{Introduction}
The \texttt{RL-world-model} implements a reinforcement learning pipeline in a quantum-inspired environment. The workflow consists of dataset generation, latent space modeling, and agent training. Below, we describe the key components:

\section*{Dataset Generation}
The script \texttt{generate\_dataset.py} performs Breadth-First Search (BFS) over a quantum state space:
\begin{itemize}
    \item States are deduplicated using global-phase invariance.
    \item Each unique state is fingerprinted as a binary vector.
    \item A transition table is generated, mapping actions to next states.
\end{itemize}
The dataset is saved in \texttt{.npz} format with the following fields:
\[
\text{fingerprints}, \text{transitions}, \text{depths}, \text{action\_names}.
\]

\section*{Latent Space Modeling}
\subsection*{MLP-VAE}
The \texttt{mlp\_vae.py} defines a Variational Autoencoder (VAE) with the following structure:
\begin{align*}
\text{Encoder:} & \quad \text{Input} \to \text{Hidden Layers} \to (\mu, \log \sigma^2) \\
\text{Decoder:} & \quad \text{Latent Vector} \to \text{Hidden Layers} \to \text{Output}.
\end{align*}
The VAE maps high-dimensional fingerprints to a low-dimensional latent space.

\subsection*{Latent Dynamics Model}
The \texttt{train\_world\_model.py} trains a model $f_v$ to predict the next latent state:
\[
f_v(\mu, a) \to \mu',
\]
where $\mu$ is the current latent state, $a$ is a one-hot encoded action, and $\mu'$ is the predicted next state.

\section*{Reinforcement Learning}
The \texttt{train\_rl\_agent.py} trains a Deep Q-Network (DQN) agent in the latent space:
\begin{itemize}
    \item \textbf{State:} $\mu \in \mathbb{R}^{\text{latent\_dim}}$.
    \item \textbf{Action:} Discrete actions from the transition table.
    \item \textbf{Reward:} $R = -\|\mu' - \mu_{\text{target}}\|_2$.
\end{itemize}
The agent learns to navigate the latent space to reach a target state.

\section*{Quantum Fingerprints}
The \texttt{multiqubit\_fingerprint.py} generates binary fingerprints for quantum states:
\begin{itemize}
    \item Sweeps rotation angles $\theta$ for each qubit.
    \item Samples $n_{\text{shots}}$ measurements in the $Z$-basis.
    \item Outputs a binary matrix of shape $(n_{\text{shots}}, n_{\text{qubits}} \times n_{\theta})$.
\end{itemize}

\section*{Conclusion}
The \texttt{RL-world-model} provides a modular framework for reinforcement learning in a quantum-inspired latent space. It combines quantum state simulation, latent space modeling, and RL agent training.

\end{document}